RELIKE: Reionization Effective Likelihood from Planck 2018 Data

### 2021-04-16 paper/rei.tex

\title{\ch{RELIKE}: Reionization Effective Likelihood from Planck 2018 Data (?)}

@@

We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested. We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual \ch{global} ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering $\tau_{\rm 2step}(15, 30) < 0.017 $ (95\% C.L.) using a  two-step toy model without using PCs.}

@@

\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}.}

@@

\ch{I'm here}
Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}

@@

\todo{write this part}
The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].

@@

Our code is available on GitHub at {give URL\url{...}}. The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.  
This is also the default analysis for all results that follow except where noted.
The most important part of the code release is the python package called RELIKE which provides the PC parameters for a model and also
% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
implements the Gaussian effective likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample RELIKE within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.

\wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
We also provide a pre-assembled CosmoMC code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.

@@

To quantify these residual oscillations we consider 
the 95\% upper limit on the high-redshift optical depth.
For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while $\tau_{\rm tanh}(15,\zmax) \approx 0$ from the exact likelihood.
Ignoring this small effect of the oscillations, 
we can see that the tanh model does not allow for high redshift ionization by construction.  As we shall see in the next example the data themselves do allow for a small but significantly larger fraction of the total optical depth to originate from high redshift, when the form of the model permits it.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}). }\todo{CH: say in text 0.026 is 2D marginalized 95\%, 0.02 is 1D.}
%best-fit (0.05290344, 0.005859431)
%95%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
\label{fig:two_step_model}
\end{figure}

@@

\begin{figure}[t]
%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
%
% longer Gaussian chains only 
%\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
%
% longer KDE and Gaussian chains
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
\caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L. contours).  The exact, KDE PC and Gaussian PC likelihoods again agree very well with a small contribution from high-$z$ through $\tau_{\rm hi}$ still allowed but not significantly favored.   Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters 
which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains coming up}
}
\label{fig:two_parameter_model_2D}
\end{figure}

@@

\ch{comment here about 2D 95\% model picked with tau hi = 0.026 is higher because 0.021 is the 1D marginalized result.}
Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$:
\ch{
\beq
\tau_{\rm 2step} = 0.0621^{+0.0062}_{-0.0061}.
\eeq
}
Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
\begin{equation}
    \tau_{2\text{step}}(15, 30) < 0.016 \quad (95\% \text{ C.L., Exact)}
    \label{eq:2stepbound}
\end{equation}
and 
$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods
Note also that the 95\% C.L. allowed two-step model in 
Fig.~\ref{fig:two_step_model} has
\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}\todo{Need to change all 95\% upper limit on tau(15,30) to a one-sided computation instead of 2.}\todo{update plot two-step with longer chain results, run 2 more days exact two-step chains.} \todo{point out in text that 95\% upper limit on tau(15,30) not the same as plot.}

@@

In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this two-step model. In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ obtained by integrating over the exact $x_e(z)$ two-step model in the exact likelihood chains, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.   
As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.  Notice again the amount of high redshift optical depth that is allowed in the two step model vs the tanh model (cf.~Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
With this in mind, we will now turn to the model-independent constraints derived from the PC analysis to generalize these results on high redshift optical depth beyond the two-step model.

@@

 We first note that by virtue of Eq.~(\ref{eq:cumtauPC}), the total optical depth receives contribution from all PCs.  However
 %
%\beq
%\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
%\eeq
%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
the first two components
\beq
\tau_{\rm PC} \approx \tau_{12} \equiv  m_1 \tau_1 + m_2 \tau_2 +
\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) 
\label{eq:tau12}
\eeq
where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of the fiducial ionization model.  We explicitly demonstrate this approximate equivalence in Fig.~\ref{fig:tau12} by comparing the
posterior distributions of $\tau_{\rm PC}$ and $\tau_{12}$ both with and without the physicality prior.

@@

More generally, to the extent that the data constrain $m_1$ and $m_2$ better than the flat range-bound priors, these priors become irrelevant and the reweighting of Ref.~\cite{Millea:2018bko} becomes erroneously motivated.   To see this, consider a range-bound prior that is rotated to be oriented along constant $\tau_{12}$ (see Fig.~\ref{fig:prior_box}, blue dashed rectangle). This rotated prior in $m_1-m_2$ space corresponds to a flat prior in $\tau_{12}$ space.
Had the original prior (blue square) been shifted to slightly lower
$m_1$ and encompassed all of the 95\% CL allowed
region the two priors would yield  identical constraints at the
95\% CL.   This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would still change the original prior but not the
rotated one.  
%will try rephrasing.}

@@

Finally, we check that our results (Eq.~\ref{eq:tau_pc_zmax30}) are robust to
 extending the PC
range to
$z_{\rm max}=50$ which is negligibly different
\ch{
\beq
%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
\tau_{\rm PC} = 0.0626^{+0.0061}_{-0.0072}\quad (\zmax=50).
\eeq
CH: removing SRoll2 label in parenthesis and moving zmax=50 into the parenthesis.}
Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
\beq
\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}).
\eeq
As with the tanh model, the \texttt{SRoll2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.

@@

The value of a model independent approach is even more apparent for constraints on the
cumulative optical depth $\tau(z, \zmax)$ at high $z$.
The main results are again related to the $m_1$ and $m_2$ PC constraints. 
As discussed in Ref.~\cite{Heinrich:2016ojb}, raising $m_1$ increases the  optical depth  at all $z$ but with a larger weighting at high $z$ whereas
raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
Consequently, as shown in Fig.~\ref{fig:prior_box}, the second aspect of PC constraints beyond the well constrained total optical depth, as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12}), is a positive
correlation between the two components that represents the high to low $z$ 
contribution to the total optical depth.  Any finer redshift detail in the ionization history is much more poorly constrained.

In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
the 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.  Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.  Moreover upper bounds on the optical depth at high 
redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example,
\ch{
\beq
\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.}).
\eeq
}
This should be contrasted with the tanh model 
 where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
 On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~\ref{eq:2stepbound}).  It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
 \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..}

@@

In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.  Improvements from srollv2 come mainly from improving the lower bound at low redshift, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.

@@

For the high-redshift optical depth, we also obtain consistent upper limits
%\beq
%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{SRoll2}),
%\eeq
%[double check on full chain (from 10/15) this is from 09/30.]
%and 
\beq
\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.}).
\eeq

@@

\begin{itemize}
    \item other interesting points to discuss and add 
    \item future outlook (mention joint analyses)
\end{itemize}

>>>>

\title{RELIKE: Reionization Effective Likelihood from Planck 2018 Data}

@@

We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested. We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual global ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as \ch{line-intensity mapping},
%21cm,
luminosity function, star formation history, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering $\tau_{\rm 2step}(15, 30) < 0.017 $ (95\% C.L.) using a  two-step toy model without using PCs.}

@@

\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}. mean free path of ionizing photons  between 5 and 6: \url{https://arxiv.org/pdf/2103.16610.pdf}, $<80\%$ ionized at z = 6. }

@@

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}

@@

The code is publicly available on GitHub at \url{https://github.com/chenheinrich/RELIKE}. We publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model. If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space; we provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in \texttt{CosmoMC-relike}. We expect the code release to dramatically reduce the time for testing individual models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity function, star formation history, 21cm or line-intensity mapping in general, etc.

@@

Our code is available on GitHub at {give URL\url{...}}. The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.  
This is also the default analysis for all results that follow except where noted.
The most important part of the code release is the python package called RELIKE which provides the PC parameters for a model and also
% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
implements the Gaussian effective likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use RELIKE to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample RELIKE within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.

We also provide a pre-assembled CosmoMC code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.

@@

To quantify these residual oscillations we consider 
the 95\% upper limit on the high-redshift optical depth.
For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while \ch{$\tau_{\rm tanh}(15,30) \approx 0$}
%$\tau_{\rm tanh}(15,\zmax) \approx 0$ 
from the exact likelihood. \ch{NEW: since both one-sided and two-sided are 0.002 here, I'm doing the one-sided vs two-sided explaination in the two-step section instead.}
Ignoring this small effect of the oscillations, 
we can see that the tanh model does not allow for high redshift ionization by construction.  As we shall see in the next example the data themselves do allow for a small but significantly larger fraction of the total optical depth to originate from high redshift, when the form of the model permits it.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}). }
%best-fit (0.05290344, 0.005859431)
%95%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
\label{fig:two_step_model}
\end{figure}

@@

\begin{figure}[t]
%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
%
% longer Gaussian chains only 
%\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
%
% longer KDE and Gaussian chains
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
\caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L. contours).  The exact, KDE PC and Gaussian PC likelihoods again agree very well with a small contribution from high-$z$ through $\tau_{\rm hi}$ still allowed but not significantly favored.   Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters 
which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains (still running)}
}
\label{fig:two_parameter_model_2D}
\end{figure}

@@

\ch{NEW: $\tau_{\rm hi}$ upper limits are all one-sided already.}\\

\ch{Note that the 95\% upper limit on $\tau_{\rm hi} < 0.021$ presented here is a 1D marginalized result; whereas the model shown in Figs~\ref{fig:two_parameter_model_2D} and~\ref{fig:two_parameter_model_2D} which has $\tauhi = 0.026$ is a model picked at the 95\% C.L. in the 2D marginalized case, hence the difference from the 1D result here.}
Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$:
\ch{
\beq
\tau_{\rm 2step} = 0.0621^{+0.0062}_{-0.0061}.
\eeq
}
Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
\begin{equation}
    \tau_{2\text{step}}(15, 30) < 0.016 \quad (95\% \text{ C.L., Exact)}
    \label{eq:2stepbound}
\end{equation}
\todo{CH: still missing one-sided $\tau_{2\text{step}}(15, 30)$ for exact.}
and 
\ch{$\tau_{2\text{step}}(15, 30) < 0.016$} for both effective likelihoods. %changed from <0.017--> <0.016 when going from two-sided to one-sided.
Note also that the 95\% C.L. allowed two-step model in 
Fig.~\ref{fig:two_step_model} has
\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}
\ch{Note that the 95\% upper limit on $\tau(15, 30)$ reported in the text are one-sided upper limits, which are slightly different than the two-sided upper limits plotted in the figures of cumulative optical depth (e.g. Fig.~\ref{fig:plot_taugtz_two_step_contours})}

@@

In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this two-step model. we show the 68\% and 95\% limits on $\tau(z, \zmax)$ obtained by integrating over the exact $x_e(z)$ two-step model in the exact likelihood chains, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.   
As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.  Notice again the amount of high redshift optical depth that is allowed in the two step model vs the tanh model (cf.~Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
With this in mind, we will now turn to the model-independent constraints derived from the PC analysis to generalize these results on high redshift optical depth beyond the two-step model.

@@

 We first note that by virtue of Eq.~(\ref{eq:cumtauPC}), the total optical depth receives contribution from all PCs.  However
 %
%\beq
%\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
%\eeq
%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
the first two components
\beq
\tau_{\rm PC} \approx \tau_{12} \equiv  m_1 \tau_1 + m_2 \tau_2 +
\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) ,
\label{eq:tau12}
\eeq
where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of the fiducial ionization model.  We explicitly demonstrate this approximate equivalence in Fig.~\ref{fig:tau12} by comparing the
posterior distributions of $\tau_{\rm PC}$ and $\tau_{12}$ both with and without the physicality prior.

@@

More generally, to the extent that the data constrain $m_1$ and $m_2$ better than the flat range-bound priors, these priors become irrelevant and the reweighting of Ref.~\cite{Millea:2018bko} becomes erroneously motivated.   To see this, consider a range-bound prior that is rotated to be oriented along constant $\tau_{12}$ (see Fig.~\ref{fig:prior_box}, blue dashed rectangle). This rotated prior in $m_1-m_2$ space corresponds to a constant $P(\tau_{12}$) once the direction orthogonal to $\tau_{12}$ is marginalized.
Had the  \wh{data constraint been shifted to slightly higher
$m_1$ so that both prior ranges encompassed all of the 95\% CL allowed
region they both would yield  the same constraints at that level.}
% rephrased to flow with the next paragraph of 2015->2018
This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would  change the original prior but not the
rotated one and \wh{destroy this desired equivalence.  More generally the point is that for a multidimensional space, constructing a flat marginal 1D distribution per parameter in the absence of data does not guarantee a locally flat prior in the region supported by the data
(see \cite{Heinrich:2018btc} for further discussion).}
%will try rephrasing.}

@@

Finally, we check that our results (Eq.~\ref{eq:tau_pc_zmax30}) are robust to
 extending the PC
range to
$z_{\rm max}=50$ which is negligibly different
%
\beq
%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
\tau_{\rm PC} = 0.0626^{+0.0061}_{-0.0072}\quad (\zmax=50).
\eeq
%
Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
\beq
\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}).
\eeq
As with the tanh model, the \texttt{SRoll2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.

@@

The value of a model independent approach is even more apparent for constraints on the
cumulative optical depth $\tau(z, \zmax)$ at high $z$.
The main results are again related to the $m_1$ and $m_2$ PC constraints. 
As discussed in Ref.~\cite{Heinrich:2016ojb}, raising $m_1$ increases the  optical depth  at all $z$ but with a larger weighting at high $z$ whereas
raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
Consequently, as shown in Fig.~\ref{fig:prior_box}, beyond the well constrained total optical depth, as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12}), orthogonal to this combination is a somewhat weaker constraint on the the high to low $z$ 
contribution to the total optical depth.  Any finer redshift detail in the ionization history is much more poorly constrained.

In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
the \ch{two-sided} 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.  Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.  Moreover upper bounds on the optical depth at high 
redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example,
\ch{
\beq
\tau_{\rm PC}(15, 30) < 0.020 \; (95\%\; \mathrm{C.L.}).
\label{eq:tauPC15}
\eeq
}
\ch{Recall again that we are reporting the one-sided upper limit here in the text, which is slightly lower than the two-sided limits shown in the figures.}
%
\ch{(NEW: changed from 0.023 to 0.020 going from two-sided to one-sided.)}
This should be contrasted with the tanh model 
 where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
 On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~\ref{eq:2stepbound}).  It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
 \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..} \ch{$\tau_{\rm hi}$ values are confirmed to be one-sided now! :) }

@@

In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.  Improvements from \texttt{SRoll2} come mainly from improving the lower bound at low redshift, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.

@@

For the cumulative high-redshift optical depth,
\beq
\tau_{\rm PC}(15, 50) < 0.019 \; (95\%\; \mathrm{C.L.}),
\eeq
we also obtain consistent upper limits to the $\zmax=30$ case of Eq.~(\ref{eq:tauPC15}).
\ch{NEW: changed 0.022 to 0.019 going from two-sided to one-sided}

@@

In the future, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f. adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects. [Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data? More importantly with data in general that wants to vary As during sampling? Put prior on As consistent with Planck?].

There has also been recent studies revealing that reionization might not have ended by $z = 6$~\cite{} [cite]. If this is true, the current PC approach will need be modified to accommodate a range $z_{\rm min} < z < z_{\rm max}$ with a a lower $z_{\rm min} < 6$. We leave this for future work as it would require constructing a new set of PC functions.

[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]

In sum, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on specific ionization models, as well as to open the way to jointly constrain legacy CMB large-scale data with a variety of increasingly rich reionization datasets coming up in the next decade.
